Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-12-04 AuthorResponseToReviewers.Rmd

*The most crucial point is in my opinion the missing discussion of the influence of different data distributions. In the visualization community exists research on different descriptors for two-dimensional data called Scagnostics. When I started reading this article, I expected a thorough discussion of different data distributions/properties and the influence on different binning strategies and error outcomes.*

- We are very grateful that the review brought the literature on scagnostics to our attention; the work and scagnostic metrics formalized in the works by Wilkerson, Anand, Grosmann and Wills will be valuable to the discussion of binned scatterplots. Specifically scagnostic literature contributes to Section 5 of our paper in the exploration of loss under a variety of simulated distribution. You will find that a discussion of the relevance of scagnostics has been added immediately preceding the start of subsection 5.1. It is interesting that the striation score is so small, given that the data resolution issue can lead to artificially induced stripping is seen in so prominently in Figure 6.

*In Section 4.2, a discussion on contrast effects hindering the perception of the encoded values is missing. Nearby colored points, will change the color impression. These effects can be modelled and estimated and maybe interesting to investigate a little further.*

- Citations to refer readers to Adelson and Fairchild as example of work on contrast effects has been added to Section 4.2 and related future work has been added to the discussion in Section 6.

@@

### Responses to Reviewer 2 Comments/Questions

*This paper reviews visualization methods for scatterplots and their strengths and weaknesses in handling large datasets. Section 2 notes that binned aggregation is the most widely used and effective way to visualize big bivariate data. In Section 3, the paper introduces three types of binning algorithms: standard rectangular, random and quantile. The main idea in this paper is presented in Section 4, where loss of information is characterized using two metrics: spatial loss and frequency loss. Spatial loss accounts for the loss of information about each individual point’s location; and frequency loss addresses the limitation of human perception in distinguishing frequencies represented using hues or saturation. Section 5 then presents some empirical results on how various factors may affect spatial loss and frequency loss based on three synthetic datasets. These factors include bin width and bin origin for spatial loss, and number of bins for frequency loss.*

*In general, I think this paper addresses an important topic. Guidelines on choosing binning algorithms will be immensely helpful for practitioners dealing with big data. I am, however, reserved in recommending acceptance because of two main concerns: lack of novelty and unconvincing results.*

*Lack of Novelty:*
*Section 2 and 3 do a good job in reviewing the concepts of binned aggregation and varieties of binning algorithms. However these are fairly well known and the discussion does not seem to contribute novel ideas. The only thing that I haven’t seen before is the random binning approach. Is this a new idea proposed by you? If so, please state clearly. I also wonder how and why random binning can be useful. Please provide scenarios in which it would make sense to use random binning. For the other ideas covered in these two sections, please properly cite relevant work.*

- You are correct that many of the ideas in Sections 2 and 3 are not original, with the exception of the random binning approach, where the methodologies for random-binned aggregation and post processing for net spatial loss are original to our best knowledge. The goal of these sections is to bring many binning related concepts under a unified notation for the benefit of clarity in the sections to follow. For the random binning, clearer language of proposal of the idea and a statement has been added to address the motivation for its introduction. For the unoriginal ideas, additional citations have been made to place our unifying notation better within the existing literature.

@@

*I do not understand how binned frequencies (as compared to true frequencies) are calculated in Section 4.2 on frequency loss. Please provide an example. Also please clarify is binned frequencies are measures of frequencies perceived by a human.*

- A clarifying example using the visual from Figure 3 has been added to Section 4.2.

*The first paragraph of Section 5.1 shows that spatial loss increases as bin size increases, this seems hardly surprising. In the second paragraph, you talk about three properties of visual estimator: unbiasedness, consistency and efficiency, I think these properties are interesting, but the rest of the paper does not address them sufficiently.*

- The interesting parallels with common properties of statistical estimators were used as a motivating concept for the construction of binning loss functions. We felt that an extended exploration of this analog would distract from the exploration on optimizing bin specification. The need for future work on exploring this topic has been added to the discussion in Section 6.

@@

- The main result from the paper is to provide simple adjustment recommendations for binning specification for reducing loss in binned scatterplots. To emphasize these recommendations, the baseball plot example in Section 5.3 has been reworked to highlight the take-home message. To your other point, we agree that future work to quantify the implicit loss of information (e.g. the imprecise ability to map continuously shaded tiles to a frequency legend) in reading binned scatterplots is needed; however, the current work strives to quantify explicit loss of information when the perception tasks are more simple (e.g. mapping distinguishably-discrete shaded tiles to an interval on the frequency legend).

### Responses to Editors Comments/Questions
*Editor's Comments to Author:*

*Associate Editor - Comments to the Author:*
*The paper in its current form does not provide a sufficiently thorough coverage of either review material or novel material to warrant publication.  However, the reviewers have some useful suggestions for ways in which a different paper (or two) could possibly be produced by expanding on different parts of this submission.*

- We hope that the revisions explanding review material and clarification of our contributions to the topic will meet the standards of your publication. If there are any additional comments, concerns or questions, our team is happy to address them.

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Expanded on color contrast impact on perception a bit. Added Adelson and Fairchild citations.

% It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates. The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from over-plotting, which can be immense for large data sets. When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location. This implicit loss of frequency information is then exacerbated when points are rendered as a circle with a radius large enough to also partially cover nearby points. In binned scatterplots, we have traded the exact locations for exact frequencies within bins. 
% 
Bin counts can be displayed using a continuous shading scale in HSV color space \citep{colorperception} and thus we can theoretically map frequency to shade perfectly. While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise. Color perception of shade is extremely context sensitive -- see Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors -- allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend. It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.  %Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.

@@

Large bivariate data sets are very difficult to visualize in raw form, due to over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system. Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in spatial and frequency information.

We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms. The random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occur when data recorded to a coarse resolution are binned using a bin width that was a non-integer multiple of the data resolution. However, the standard binning algorithm is superior due to lower spatial loss. For data with a coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) artificial stripes in the standard binning process can be avoided, if  bin dimensions are chosen as integer multiples of $\alpha_x$ and $\alpha_y$.  We were also able to show through simulation that a reasonable default for the binning uses an origin offset of $(\alpha_x/2, \alpha_y/2)$ because it resulted in minimal or near minimal spatial loss for symmetric data and performed well for heavily skewed data.

@@

\ktm{Due to imperfect human perception may elect to use frequency binning to discretize the shading scale with the goal to make the loss of frequency information more explicit. If we aim to have highly accurate perceptitual mapping,} it is recommended to use at most between four and seven distinct shades. This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts. \ktm{Frequency binning to discretize the shading scale allows us to explicitly quantify the loss of frequency information, however future perception research is needed to quantify the implicit frequency loss in reading binned scatterplots usign continuous shades.}

>>>>

% It is important to note that there is no spatial loss for the traditional scatterplot, as the points are rendered at exact coordinates. The reason we are willing to sacrifice precision of location information is to alleviate the inherent loss of frequency information from over-plotting, which can be immense for large data sets. When a point is rendered in a traditional scatterplot there is no graphical change once any additional points are placed at the same coordinates; thus visually implying a frequency of one data value at the location. This implicit loss of frequency information is then exacerbated when points are rendered as a circle with a radius large enough to also partially cover nearby points. In binned scatterplots, we have traded the exact locations for exact frequencies within bins. 
% 
Bin counts can be displayed using a continuous shading scale in HSV color space \citep{colorperception} and thus we can theoretically map frequency to shade perfectly. While the tiles of the binned scatterplot would be \textit{rendered} precisely, the ability of a human with average vision to extract that information by visually mapping the tile shade to a frequency scale in the plot legend is largely imprecise. Color perception of shade is \ktm{influenced by the contrasting effect of surrounding shades}, allowing an inaccurate mapping of tile shade to the corresponding shade in the plot legend \ktm{\citep{adelson1993perceptual,fairchild2013color}}. See Figure~\ref{fig:FreqLossDemoPlot} for a demonstration of context sensitivity of colors. It is therefore not realistic, to expect readers to be able to decode frequency from a continuous shade scheme, even though theoretically we can perceive shades continuously \citep{Leong}.  %Even though there is a lot of spread between the colors in this example, most onlookers have trouble answering the question about the relative relationship of the shading of the darkest/lightest points in each cloud of points.

@@

Large bivariate data sets are very difficult to visualize in raw form, \ktm{with the implicit loss of information due to} over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system. Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in spatial and frequency information. \ktm{We have introduced spatial and frequency loss functions to explicitly quantify the lost information in the construction of the binned scatterplot. This offers an optimizing criterion to guide binning specification and a foundation for future work on properties of a binned scatterplot as a visual estimator of the bivariate density -- such as unbiasedness, consistency and efficiency.}

We have presented two algorithms for spatially binning data points; standard and random rectangular binning algorithms. \ktm{Our proposed} random binning algorithm displayed strong advantage of avoiding the problem of artificial stripes that occur when data recorded to a coarse resolution are binned using a bin width that was a non-integer multiple of the data resolution. However, the standard binning algorithm is superior due to lower spatial loss. For data with a coarse resolution ($\alpha_x$ units in the X dimension and $\alpha_y$ units in the Y dimension) artificial stripes in the standard binning process can be avoided, if  bin dimensions are chosen as integer multiples of $\alpha_x$ and $\alpha_y$.  We were also able to show through simulation that a reasonable default for the binning uses an origin offset of $(\alpha_x/2, \alpha_y/2)$ because it resulted in minimal or near minimal spatial loss for symmetric data and performed well for heavily skewed data.

@@

\ktm{Due to imperfect human perception may elect to use frequency binning to discretize the shading scale with the goal to make the loss of frequency information more explicit. If we aim to have highly accurate perceptual mapping,} it is recommended to use at most between four and seven distinct shades. This is done to minimize the frequency loss within the bounds of human perceptual ability to distinguish multiple shades simultaneously. Using quantile frequency binning or standard log frequency binning are shown to be reasonable methods -- with slight differences of interpretability -- for handling situations with heavily skewed bin counts. \ktm{Frequency binning to discretize the shading scale allows us to explicitly quantify the loss of frequency information, however future perception research is needed to quantify the implicit frequency loss in reading binned scatterplots using continuous shading and the impact of contrasting effects.}
